---
title: "Community Responses to SWISLR - Community Data Processing and Analysis"
author: "Aeran Coughlin"
date: "2026-02-19"
output:
  html_document:
    toc: true
---

```{r}
# Install packages
library(tidyverse)
library(lme4)
library(ape)
library(car)
library(gdm)
library(vegan)
library(labdsv)
library(reshape2)
library(ggpubr)
library(TITAN2)

# Load files
env <- read.csv("swislr_env.csv") # soil and geospatial data
cover <- read.csv("swislr_cover.csv") # understory vegetation
pcq <- read.csv("swislr_pcq.csv") # trees and shrubs
spec <- read.csv("swislr_veg-species.csv") # list of all plant species
fun.otu <- read.csv("swislr_fungal-table.csv") # fungal OTU table
fun.tax <- read.csv("swislr_fungal-taxa.csv") # fungal taxonomy table
bac.otu <- read.csv("swislr_bac-table.csv") # bacterial OTU table
bac.tax <- read.csv("swislr_bac-taxa.csv") # bacterial taxonomy table

# Set color scheme for plots
color <- c("Woody Vegetation" = "#009C29", 
           "Fungal" = "#FF7CAC", 
           "Understory"= "#E8E800", 
           "Bacterial"="#2949C5", 
           "Dead Tree" = "#2F2F2F", 
           "Shrub" = "#abab02")
```

### Data Processing and Reformatting

##### Trees and Shrubs

We collected circumference of overstory species and diameter of understory species. We calculate basal area for both types and then merge into a single column, then basal area is summed by species for each plot.

```{r}
# Calculate basal area 
# A = c^2 / 4*pi
# A = pi*d^2 / 4
pcq.sum <- pcq %>% 
  mutate(over.ba = ((circumference^2) / (4*3.141593))) %>% # overstory (from circumference)
  mutate(under.ba = ((3.141593*(diameter^2)) / 4)) %>% # understory (from diameter)
  mutate(ba = coalesce(over.ba, under.ba)) %>% # merge over and understory
  select(plot.code, species, ba) %>% 
  group_by(plot.code, species) %>% 
  subset(species!="") %>% 
  summarise(ba.sum=sum(ba)) %>% # sum basal area by plot and species
  as.data.frame() %>%
  arrange(plot.code)

# Convert df to species matrix for analysis
ba.matrix <- matrify(pcq.sum)

ba.long <- melt(as.matrix(ba.matrix))

ba.sum <- ba.long %>% 
  rename("plot.code"="Var1") %>% 
  rename("species"="Var2") %>% 
  rename("ba.sum"="value")
```


##### Understory Vegetation

Understory species abundances were collected as percent cover.

```{r}
## convert all values to numeric ##
cover.sum <- cover %>% 
  select(-c(year)) %>% 
  mutate(percent.cover = as.character(percent.cover)) %>% 
  mutate(percent.cover = replace(percent.cover, percent.cover == "<1", "0.5")) %>% 
  mutate(percent.cover = as.numeric(percent.cover))

# Convert df to species matrix for analysis
cov.matrix <- matrify(cover.sum)
```

##### Fungi

```{r}
# Assess read numbers across samples
fun.otu %>% 
  mutate(rowsum = rowSums(select_if(., is.numeric), na.rm = TRUE)) %>% 
  select(plot.code, rowsum) %>% 
  summarise(min=min(rowsum))

# Change plot code column to row names
fun.matrix <- fun.otu %>%
  column_to_rownames(var="plot.code")
```


##### Bacteria

```{r}
# Assess read numbers across samples
bac.otu %>% 
  mutate(rowsum = rowSums(select_if(., is.numeric), na.rm = TRUE)) %>% 
  select(plot.code, rowsum) %>% 
  summarise(min=min(rowsum)) #12684

# Change plot code column to row names
bac.matrix <- bac.otu %>% 
  column_to_rownames(var="plot.code")
```


### Vegetation Abundance Analysis

Testing hypotheses about which SWISLR stressor best predicts the abundance of vegetation.

```{r}
# Create one df of abundance of all vegetation types

# Herbaceous understory percent cover 
herb.sum <- cover.sum %>% 
  left_join(spec, by="species") %>% 
  filter(category!="tree" & category!="shrub") %>%
  group_by(plot.code) %>% 
  summarise(herb.sum=sum(percent.cover)) %>% 
  mutate(herb.sum=replace(herb.sum, herb.sum > 100, 100))

# Shrub basal area total
ba.sum.shrub <- ba.sum %>% 
  filter(species=="Baccharis halimifolia" | 
           species=="Clethra alnifolia" | 
           species=="Ilex coriacea" | 
           species=="Ilex glabra" | 
           species=="Lyonia lucida" | 
           species=="Myrica cerifera" |
           species=="Persea palustris" |
           species=="Vaccinium sp.") %>% 
  group_by(plot.code) %>% 
  summarise(shrub.ba=sum(ba.sum))

# Tree basal area log total and presence/absence
ba.log.sum.tree <- ba.sum %>% 
  mutate(log.ba.sum = if_else(ba.sum > 0, log(ba.sum), 0)) %>% 
  filter(species=="Pinus taeda" | 
           species=="Liquidambar styraciflua" | 
           species=="Acer rubrum" | 
           species=="Pinus serotina") %>%
  group_by(plot.code) %>% 
  summarise(tree.log.ba=sum(log.ba.sum)) %>% 
  mutate(tree.pres = if_else(tree.log.ba > 0, 1, 0))

# Tree basal area total
ba.sum.tree <- ba.sum %>% 
  filter(species=="Pinus taeda" | 
           species=="Liquidambar styraciflua" | 
           species=="Acer rubrum" | 
           species=="Pinus serotina") %>%
  group_by(plot.code) %>% 
  summarise(tree.ba=sum(ba.sum))

# Dead tree area total
ba.sum.dead <- ba.sum %>% 
  filter(species=="dead tree") %>% 
  group_by(plot.code) %>% 
  summarise(dead.sum=sum(ba.sum)) 

# Combine
veg.ba.sum <- env %>% 
  left_join(herb.sum, by="plot.code") %>% 
  left_join(ba.sum.shrub, by="plot.code") %>% 
  left_join(ba.log.sum.tree, by="plot.code") %>% 
  left_join(ba.sum.tree, by="plot.code") %>% 
  left_join(ba.sum.dead, by="plot.code")

veg.ba.sum[is.na(veg.ba.sum)] <- 0 # for plot with no woody vegetation
```

##### Understory Vegetation

```{r}
# Hypotheses
herb.h1 <- lmer(herb.sum~ppt + (1|transect), data=veg.ba.sum)
herb.h2 <- lmer(herb.sum~elevation + (1|transect), data=veg.ba.sum)
herb.h3 <- lmer(herb.sum~avg.pH + (1|transect), data=veg.ba.sum)
herb.h4 <- lmer(herb.sum~ppt + elevation + (1|transect), data=veg.ba.sum)
herb.h5 <- lmer(herb.sum~ppt + avg.pH + (1|transect), data=veg.ba.sum)
herb.h6 <- lmer(herb.sum~avg.pH + elevation + (1|transect), data=veg.ba.sum)
herb.h7 <- lmer(herb.sum~ppt + elevation + avg.pH + (1|transect), data=veg.ba.sum)

AIC(herb.h1, herb.h2, herb.h3, herb.h4, herb.h5, herb.h6, herb.h7)
Anova(herb.h7)
```

##### Shrubs

```{r}
# Hypotheses
shrub.h1 <- lmer(shrub.ba~ppt + (1|transect), data=veg.ba.sum)
shrub.h2 <- lmer(shrub.ba~elevation + (1|transect), data=veg.ba.sum)
shrub.h3 <- lmer(shrub.ba~avg.pH + (1|transect), data=veg.ba.sum)
shrub.h4 <- lmer(shrub.ba~ppt + elevation + (1|transect), data=veg.ba.sum)
shrub.h5 <- lmer(shrub.ba~ppt + avg.pH + (1|transect), data=veg.ba.sum)
shrub.h6 <- lmer(shrub.ba~avg.pH + elevation + (1|transect), data=veg.ba.sum)
shrub.h7 <- lmer(shrub.ba~ppt + elevation + avg.pH + (1|transect), data=veg.ba.sum)

AIC(shrub.h1, shrub.h2, shrub.h3, shrub.h4, shrub.h5, shrub.h6, shrub.h7)
Anova(shrub.h7)
```

##### Trees

```{r}
# Two step model
# Step 1: presence-absence 
pa.h1 <- glm(tree.pres~elevation,  family=binomial(link="logit"), data=veg.ba.sum)
pa.h2 <- glm(tree.pres~ppt,  family=binomial(link="logit"), data=veg.ba.sum)
pa.h3 <- glm(tree.pres~avg.pH,  family=binomial(link="logit"), data=veg.ba.sum)
pa.h4 <- glm(tree.pres~ppt + elevation,  family=binomial(link="logit"), data=veg.ba.sum)
pa.h5 <- glm(tree.pres~ppt + avg.pH,  family=binomial(link="logit"), data=veg.ba.sum)
pa.h6 <- glm(tree.pres~elevation + avg.pH,  family=binomial(link="logit"), data=veg.ba.sum)
pa.h7 <- glm(tree.pres~ppt + elevation + avg.pH,  family=binomial(link="logit"), data=veg.ba.sum)

AIC(pa.h1, pa.h2, pa.h3, pa.h4, pa.h5, pa.h6, pa.h7)
Anova(pa.h2)

# Step 2: plots with 0 tree ba removed
# Hypotheses
tree.h1 <- lmer(tree.log.ba~ppt + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))
tree.h2 <- lmer(tree.log.ba~elevation + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))
tree.h3 <- lmer(tree.log.ba~avg.pH + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))
tree.h4 <- lmer(tree.log.ba~ppt + elevation + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))
tree.h5 <- lmer(tree.log.ba~ppt + avg.pH + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))
tree.h6 <- lmer(tree.log.ba~avg.pH + elevation + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))
tree.h7 <- lmer(tree.log.ba~ppt + elevation + avg.pH + (1|transect), data=veg.ba.sum %>% filter(tree.pres==1))

AIC(tree.h1, tree.h2, tree.h3, tree.h4, tree.h5, tree.h6, tree.h7)
Anova(tree.h6)
```


### Multidimensional Scaling (MDS) 

##### Trees and Shrubs

```{r}
# Relativize basal area
pcq.matrix.rel <- decostand(ba.matrix, method = "total")

# Calculate Bray-Curtis Dissimilarity  
pcq.matrix.dist <- vegdist(pcq.matrix.rel, method = "bray")

# Run MDS
pcq.mds <- metaMDS(pcq.matrix.dist, distance = "bray", trymax=1000)

# Assess goodness of fit with Shepard Diagram
stressplot(pcq.mds)

# Extract MDS axis values 
pcq.mds.scores <-  as.data.frame(pcq.mds$points)

pcq.mds.scores.plot <- pcq.mds.scores %>% 
  rownames_to_column(var="plot.code") %>% 
  left_join(env, by="plot.code")

```

##### Understory Vegetation

```{r}
# Relativize percent cover
cov.matrix.rel <- decostand(cov.matrix, method = "total")

# Calculate Bray-Curtis Dissimilarity 
cov.matrix.dist <- vegdist(cov.matrix.rel, method = "bray")

# Run MDS
cov.mds <- metaMDS(cov.matrix.dist, distance = "bray", trymax=1000)

# Assess goodness of fit with Shepard Diagram
stressplot(cov.mds)

# Extract MDS axis values 
cov.mds.scores <-  as.data.frame(cov.mds$points)

cov.mds.scores.plot <- cov.mds.scores %>% 
  rownames_to_column(var="plot.code") %>% 
  left_join(env, by="plot.code")
```

##### Fungi 

```{r}
# Relativize read abundance 
fun.matrix.rel <- decostand(fun.matrix, method = "total")

# Calculate Bray-Curtis Dissimilarity 
fun.matrix.dist <- vegdist(fun.matrix.rel, method = "bray")

# Run MDS
set.seed(1)
fun.mds <- metaMDS(fun.matrix.dist, distance = "bray", trymax=1000)

# Assess goodness of fit with Shepard Diagram
stressplot(fun.mds)

# Extract MDS axis values 
fun.mds.scores <-  as.data.frame(fun.mds$points)

fun.mds.scores.plot <- fun.mds.scores %>% 
  rownames_to_column(var="plot.code") %>% 
  left_join(env, by="plot.code")
```

##### Bacteria

```{r}
# Relativize read abundance 
bac.matrix.rel <- decostand(bac.matrix, method = "total")

# Calculate Bray-Curtis Dissimilarity 
bac.matrix.dist <- vegdist(bac.matrix.rel, method = "bray")

# Run MDS
bac.mds <- metaMDS(bac.matrix.dist, distance = "bray", trymax=1000)

# Assess goodness of fit with Shepard Diagram
stressplot(bac.mds)

# Extract MDS axis values 
bac.mds.scores <-  as.data.frame(bac.mds$points)

bac.mds.scores.plot <- bac.mds.scores %>% 
  rownames_to_column(var="plot.code") %>% 
  left_join(env, by="plot.code")
```

##### FIGURE 1B: NMDS Ordination Plots

```{r}
shapes <- c(21, 24, 22)
shading <- c(0.2, 0.4, 1)

cov.nmds <- ggplot(cov.mds.scores.plot, aes(x=MDS1, y=MDS2)) + 
  geom_point(aes(alpha=transect, shape=transect, fill="Understory"), size=4) + 
  geom_point(aes(shape=transect), size=4, color="black") + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  scale_shape_manual(values=shapes) +
  scale_alpha_manual(values=shading) + 
  stat_ellipse(aes(alpha=transect, color="Understory")) + 
  theme_classic() + 
  theme(legend.position="NONE") + 
  ggtitle("Understory") + 
  geom_label(aes(x = 1.5, y = 1.2, label = "R2: 0.162\nP-value: 0.019"), size=2.5)

pcq.nmds <- ggplot(pcq.mds.scores.plot, aes(x=MDS1, y=MDS2)) + 
  geom_point(aes(alpha=transect, shape=transect, fill="Woody Vegetation"), size=4) + 
  geom_point(aes(shape=transect), size=4, color="black") + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  scale_shape_manual(values=shapes) + 
  scale_alpha_manual(values=shading) + 
  stat_ellipse(aes(alpha=transect, color="Woody Vegetation")) + 
  theme_classic() + 
  theme(legend.position="NONE") + 
  ggtitle("Woody Vegetation") + 
  geom_label(aes(x = 2.25, y = 2, label = "R2: 0.175\nP-value: 0.02"), size=2.5)

fun.nmds <- ggplot(fun.mds.scores.plot, aes(x=MDS1, y=MDS2)) + 
  geom_point(aes(alpha=transect, shape=transect, fill="Fungal"), size=4) + 
  geom_point(aes(shape=transect), size=4, color="black") + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  scale_shape_manual(values=shapes) + 
  scale_alpha_manual(values=shading) + 
  stat_ellipse(aes(alpha=transect, color="Fungal")) + 
  theme_classic() + 
  theme(legend.position="NONE") + 
  ggtitle("Fungi") + 
  geom_label(aes(x = 1.5, y = 1.25, label = "R2: 0.154\nP-value: 0.001"), size=2.5)

bac.nmds <- ggplot(bac.mds.scores.plot, aes(x=MDS1, y=MDS2)) + 
  geom_point(aes(alpha=transect, shape=transect, fill="Bacterial"), size=4) + 
  geom_point(aes(shape=transect), size=4, color="black") + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  scale_shape_manual(values=shapes) + 
  scale_alpha_manual(values=shading) + 
  stat_ellipse(aes(alpha=transect, color="Bacterial")) + 
  theme_classic() + 
  theme(legend.position="NONE") + 
  ggtitle("Bacteria") + 
  geom_label(aes(x = 2, y = 0.9, label = "R2: 0.187\nP-value: 0.002"), size=2.5)

# create common legend #
for.nmds.leg <- ggplot(pcq.mds.scores.plot, aes(x=MDS1, y=MDS2)) + 
  geom_point(aes(alpha=transect, shape=transect), size=5) + 
  stat_ellipse(aes(alpha=transect)) + 
  theme_classic() + 
  theme(legend.position="bottom")

# extract and view legend #
nmds.leg <- get_legend(for.nmds.leg)
as_ggplot(nmds.leg)

# arrange plots and save #
ggpubr::ggarrange(cov.nmds, pcq.nmds, fun.nmds, bac.nmds, nrow=2, ncol=2, legend.grob=nmds.leg, legend="bottom", labels=c("B","C","D","E"))
ggsave("Figure 1B NMDS.png", width = 6, height = 6, dpi=600)

```

### Diversity 

##### Trees and Shrubs

```{r}
# Create diversity data frame
pcq.div <- env %>% 
  arrange(plot.code) %>% 
  subset(plot.code!="PAM01") # no trees or shrubs present in this plot

# Calculate Shannon Diversity 
pcq.div$shannon <- diversity(ba.matrix, index="shannon")

# Calculate Richness
pcq.div$richness <- specnumber(ba.matrix)

# Hypotheses 
pcq.h1 <- lm(richness~ppt, pcq.div)
pcq.h2 <- lm(richness~elevation, pcq.div)
pcq.h3 <- lm(richness~avg.pH, pcq.div)
pcq.h4 <- lm(richness~ppt * elevation, pcq.div)
pcq.h5 <- lm(richness~avg.pH + elevation, pcq.div)
pcq.h6 <- lm(richness~ppt + avg.pH, pcq.div)
pcq.h7 <- lm(richness~ppt * elevation * avg.pH, pcq.div)

AIC(pcq.h1, pcq.h2, pcq.h3, pcq.h4, pcq.h5, pcq.h6, pcq.h7)
Anova(pcq.h3)
```

##### Understory Vegetation

```{r}
# Create diversity data frame
cov.div <- env %>% 
  arrange(plot.code)

# Calculate Shannon Diversity 
cov.div$shannon <- diversity(cov.matrix, index="shannon")

# Calculate Richness
cov.div$richness <- specnumber(cov.matrix)

# Hypotheses 
cov.h1 <- lm(richness~ppt, cov.div)
cov.h2 <- lm(richness~elevation, cov.div)
cov.h3 <- lm(richness~avg.pH, cov.div)
cov.h4 <- lm(richness~ppt + elevation, cov.div)
cov.h5 <- lm(richness~avg.pH + elevation, cov.div)
cov.h6 <- lm(richness~ppt + avg.pH, cov.div)
cov.h7 <- lm(richness~ppt + elevation + avg.pH, cov.div)

AIC(cov.h1, cov.h2, cov.h3, cov.h4, cov.h5, cov.h6, cov.h7)
Anova(cov.h3)
```

##### Fungi

```{r}
# Create diversity data frame
fun.div <- env %>% 
  arrange(plot.code)

# Calculate Shannon Diversity 
fun.div$shannon <- diversity(fun.matrix, index="shannon")

# Calculate Richness
fun.div$richness <- specnumber(fun.matrix)

# Hypotheses
```

##### Bacteria

```{r}
# Create diversity data frame
bac.div <- env %>% 
  arrange(plot.code)

# Calculate Shannon Diversity 
bac.div$shannon <- diversity(bac.matrix, index="shannon")

# Calculate Richness
bac.div$richness <- specnumber(bac.matrix)

# Hypotheses

```

##### FIGURE 3: Richness

```{r}
cov.div.pH <- ggplot(data=cov.div, aes(x=avg.pH, y=richness, color="Understory", fill="Understory")) + theme_classic() +
  geom_point(size=3) + 
  #geom_smooth(aes(color="Understory"), fullrange=TRUE, se=FALSE,
             # method="lm", linewidth=0.5) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  geom_smooth(fullrange=TRUE, se=TRUE,
              method="lm", linewidth=0.5, linetype="dashed") + 
  xlab("pH") +
  ggtitle(bquote(bold('A ') ~ 'Understory')) +
  theme(legend.position="NONE", 
        axis.title.y = element_blank()) + 
  geom_label(aes(x = 5.8, y = 11.5, label = "P=0.090\nR2=0.07"), color="black", fill="white", size=3)

pcq.div.pH <- ggplot(data=pcq.div, aes(x=avg.pH, y=richness, color="Woody Vegetation", fill="Woody Vegetation")) + 
  theme_classic() +
  geom_point(size=3) + 
  geom_smooth(fullrange=TRUE, se=TRUE,
              method="lm", linewidth=0.5) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  xlab("pH") +
  ggtitle(bquote(bold('B ') ~ 'Woody Vegetation')) +
  theme(legend.position="NONE", 
        axis.title.y = element_blank()) + 
  geom_label(aes(x = 5.8, y = 5.5, label = "P=0.007\nR2=0.22"), color="black", fill="white", size=3)

bac.div.pH <- ggplot(data=bac.div, aes(x=avg.pH, y=richness, color="Bacterial", fill="Bacterial")) + theme_classic() +
  geom_point(size=3,) + 
  geom_smooth(fullrange=TRUE, se=TRUE,
              method="lm", linewidth=0.5) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  xlab("pH") +
  ggtitle(bquote(bold('D ') ~ 'Bacterial')) +
  theme(legend.position="NONE", 
        axis.title.y = element_blank()) + 
  geom_label(aes(x = 5.8, y = 2575, label = "P=0.034\nR2=0.13"), color="black", fill="white", size=3)

fun.div.ppt <- ggplot(data=fun.div, aes(x=ppt, y=richness, color="Fungal", fill="Fungal")) + 
  theme_classic() +
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  geom_point(size=3) + 
  geom_smooth(fullrange=TRUE, se=TRUE,
              method="lm", linewidth=0.5, linetype="dashed") + 
  xlab("Soil salinity (ppt)") +
  ggtitle(bquote(bold('C ') ~ 'Fungal')) +
  theme(legend.position="NONE", 
        axis.title.y = element_blank()) + 
  geom_label(aes(x = 11.75, y = 630, label = "P=0.445\nR2=-0.02"), color="black", fill="white", size=3)

fig3 <- ggarrange(cov.div.pH, pcq.div.pH, fun.div.ppt, bac.div.pH, nrow=2, ncol=2)

ggsave("Figure 3 Richness.png", plot=fig3, width = 6, height = 5, dpi=600)
```


### Testing SWISLR Variable Colinearity

```{r}
# Variance inflation factor 
# Vegetation abundance models
vif(herb.h7)
vif(shrub.h7)
vif(tree.h7)

# Richness models
vif(lm(richness ~ ppt + avg.pH + elevation, data=pcq.div))
vif(lm(richness ~ ppt + avg.pH + elevation, data=cov.div))
vif(lm(richness ~ ppt + avg.pH + elevation, data=fun.div))
vif(lm(richness ~ ppt + avg.pH + elevation, data=bac.div))

# Variable correlation
cor(env$ppt, env$elevation) # -0.04868284
cor(env$elevation, env$avg.pH) # -0.2480433
cor(env$ppt, env$avg.pH) # 0.6468339
cor(env$elevation, env$percent.sm) # -0.5699092
cor(env$ppt, env$percent.sm)# 0.2215169
```

### Generalized Dissimilarity Models

```{r}
# Select relevant environmental and geographic variables to include in models
soil.gdm <- env %>% 
  dplyr::select(plot.code, ppt, avg.pH, elevation)

site.gdm <- env %>% 
  dplyr::select(plot.code, latitude, longitude) %>% 
  arrange(plot.code)
```

##### Trees and Shrubs

```{r}
# Join environmental data with pcq data in long format
pcq.xy <- ba.sum %>% 
  left_join(site.gdm, by="plot.code") %>% 
  left_join(soil.gdm, by="plot.code")

# Select species data
pcq.sppTab <- pcq.xy %>%
  dplyr::select(plot.code, longitude, latitude, species, ba.sum)

# Select environmental data
pcq.envTab <- pcq.xy %>%
  dplyr::select(plot.code, ppt, avg.pH, elevation)

pcq.xy$species <- as.factor(pcq.xy$species)
pcq.xy$plot.code <- as.factor(pcq.xy$plot.code)


# Format site pairs for GDM
pcq.gdmTab <- gdm::formatsitepair(bioData=pcq.sppTab, bioFormat=2, abundance=TRUE, dist="bray", 
                         siteColumn="plot.code", XColumn="longitude", YColumn="latitude", 
                         sppColumn="species", abundColumn="ba.sum", predData=pcq.envTab)

# Run GDM
set.seed(1)
pcq.gdm <- gdm(pcq.gdmTab, geo=TRUE)

# Model results summary
summary(pcq.gdm)

# Quantify predictor importance 
set.seed(1)
gdm.varImp(pcq.gdmTab, geo=TRUE)

# Extract model iSplines for plotting
pcq.splines <- as.data.frame(isplineExtract(pcq.gdm))

# Identify positions of steepest slope
pcq.slopes <- pcq.splines %>% 
  mutate(ppt.slope = (y.ppt - lag(y.ppt)) / (x.ppt - lag(x.ppt))) %>% 
  mutate(pH.slope = (y.avg.pH - lag(y.avg.pH)) / (x.avg.pH - lag(x.avg.pH))) %>% 
  mutate(elev.slope = (y.elevation - lag(y.elevation)) / (x.elevation - lag(x.elevation)))

# Visually assess
pcq.slopes
```

##### Understory

```{r}
# Join environmental data with cover data in long format
cov.xy <- cover %>% 
  left_join(site.gdm, by="plot.code") %>% 
  left_join(soil.gdm, by="plot.code") %>% 
  mutate(species = as.factor(species)) %>% 
  mutate(plot.code = as.factor(plot.code)) %>% 
  mutate(percent.cover = replace(percent.cover, percent.cover == "<1", "0.5")) %>% 
  mutate(percent.cover = as.numeric(percent.cover))

# Select species data
cov.sppTab <- cov.xy %>% 
  dplyr::select(plot.code, longitude, latitude, species, percent.cover)

# Select environmental data
cov.envTab <- cov.xy %>% 
  dplyr::select(plot.code, ppt, avg.pH, elevation)

# Format site pairs for GDM
cov.gdmTab <- gdm::formatsitepair(bioData=cov.sppTab, bioFormat=2, abundance=TRUE, dist="bray", 
                         siteColumn="plot.code", XColumn="longitude", YColumn="latitude", 
                         sppColumn="species", abundColumn="percent.cover", predData=cov.envTab)

# Run GDM
set.seed(1)
cov.gdm <- gdm(cov.gdmTab, geo=TRUE)

# Model results summary
summary(cov.gdm)

# Quantify predictor importance 
set.seed(1)
gdm.varImp(cov.gdmTab, geo=TRUE)

# Extract model iSplines for plotting
cov.splines <- as.data.frame(isplineExtract(cov.gdm))

# Identify positions of steepest slope
cov.slopes <- cov.splines %>% 
  mutate(ppt.slope = (y.ppt - lag(y.ppt)) / (x.ppt - lag(x.ppt))) %>% 
  mutate(pH.slope = (y.avg.pH - lag(y.avg.pH)) / (x.avg.pH - lag(x.avg.pH))) %>% 
  mutate(elev.slope = (y.elevation - lag(y.elevation)) / (x.elevation - lag(x.elevation)))

# Visually assess
cov.slopes
```

##### Fungi 

```{r}
## Create xy df of otu data ##
fun.xy <- fun.otu %>% 
  select(-c(X)) %>% 
  left_join(site.gdm, by="plot.code") %>% 
  left_join(soil.gdm, by="plot.code") %>% 
  gather(otu, n, 2:4410) %>% 
  mutate(otu = as.factor(otu)) %>% 
  mutate(plot.code = as.factor(plot.code))  

## Subset bio data ##
fun.sppTab <- fun.xy %>% 
  dplyr::select(plot.code, longitude, latitude, otu, n)

## Subset pred data ##
fun.envTab <- fun.xy %>% 
  dplyr::select(plot.code, ppt, avg.pH, elevation)

## Format site pair for GDM
fun.gdmTab <- formatsitepair(bioData=fun.sppTab, bioFormat=2, abundance=TRUE, dist="bray", 
                             siteColumn="plot.code", XColumn="longitude", YColumn="latitude", 
                             sppColumn="otu", abundColumn="n", predData=fun.envTab)

# Run GDM
set.seed(1)
fun.gdm <- gdm(fun.gdmTab, geo=TRUE)

# Model results summary
summary(fun.gdm)

# Quantify predictor importance 
set.seed(4)
gdm.varImp(fun.gdmTab, geo=TRUE)

# Extract model iSplines for plotting
fun.splines <- as.data.frame(isplineExtract(fun.gdm))

# Identify positions of steepest slope
fun.slopes <- fun.splines %>% 
  mutate(ppt.slope = (y.ppt - lag(y.ppt)) / (x.ppt - lag(x.ppt))) %>% 
  mutate(pH.slope = (y.avg.pH - lag(y.avg.pH)) / (x.avg.pH - lag(x.avg.pH))) %>% 
  mutate(elev.slope = (y.elevation - lag(y.elevation)) / (x.elevation - lag(x.elevation)))

# Assess
fun.slopes
```

##### Bacteria

```{r}
## Create xy df of otu data ##
bac.xy <- site.gdm %>% 
  left_join(bac.otu, by="plot.code") %>%
  gather(otu, n, 4:11498)

## Join xy with selected soil data ##
bac.xy.all <- bac.xy %>% 
  left_join(soil.gdm, by="plot.code")
  
## Correct column formats ##
bac.xy.all$species <- as.factor(bac.xy.all$otu)
bac.xy.all$plot.code <- as.factor(bac.xy.all$plot.code)

## subset bioData ##
bac.sppTab <- bac.xy.all %>% 
  dplyr::select(plot.code, longitude, latitude, otu, n)

## subset predData ##
bac.envTab <- bac.xy.all %>% 
  dplyr::select(plot.code, ppt, avg.pH, elevation)

## format site pair for GDM
bac.gdmTab <- formatsitepair(bioData=bac.sppTab, bioFormat=2, abundance=TRUE, dist="bray", 
                             siteColumn="plot.code", XColumn="longitude", YColumn="latitude", 
                             sppColumn="otu", abundColumn="n", predData=bac.envTab)

# Run GDM
set.seed(1)
bac.gdm <- gdm(bac.gdmTab, geo=TRUE)

# Model results summary
summary(bac.gdm)

# Quantify predictor importance 
set.seed(1)
gdm.varImp(bac.gdmTab, geo=TRUE)

# Extract model iSplines for plotting
bac.splines <- as.data.frame(isplineExtract(bac.gdm))

# Identify positions of steepest slope
bac.slopes <- bac.splines %>% 
  mutate(ppt.slope = (y.ppt - lag(y.ppt)) / (x.ppt - lag(x.ppt))) %>% 
  mutate(pH.slope = (y.avg.pH - lag(y.avg.pH)) / (x.avg.pH - lag(x.avg.pH))) %>% 
  mutate(elev.slope = (y.elevation - lag(y.elevation)) / (x.elevation - lag(x.elevation)))

# Assess
bac.slopes
```

```{r}
## FIGURE 2: Generalized Dissimilarity Model Splines ## 

salt.iS <- ggplot() + 
  geom_line(data=cov.splines, aes(x=x.ppt, y=y.ppt, color="Understory"), linewidth=1.5, linetype="dashed") + 
  geom_vline(aes(xintercept=0.05, color="Understory")) + 
  geom_vline(aes(xintercept=0.1, color="Woody Vegetation")) +
  geom_vline(aes(xintercept=0.15, color="Bacterial")) + 
  geom_vline(aes(xintercept=0.2, color="Fungal")) + 
  geom_line(data=pcq.splines, aes(x=x.ppt, y=y.ppt, color="Woody Vegetation"), linewidth=1.5) + 
  geom_line(data=bac.splines, aes(x=x.ppt, y=y.ppt, color="Bacterial"), linewidth=1.5) +
  geom_line(data=fun.splines, aes(x=x.ppt, y=y.ppt, color="Fungal"), linewidth=1.5) + 
  ylab("Ecological distance") + 
  xlab("Soil salinity (ppt)") + 
  theme_classic() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(colour = "black", linewidth=1), 
        legend.position = "NONE")  + 
  ylim(0,3.4) + 
  ggtitle("") + 
  scale_color_manual(values=color)

pH.iS <- ggplot() + 
  geom_vline(aes(xintercept=6.08, color="Understory")) + 
  geom_vline(aes(xintercept=6.09, color="Woody Vegetation")) +
  geom_vline(aes(xintercept=3.79, color="Bacterial")) + 
  geom_vline(aes(xintercept=4.53, color="Fungal")) + 
  geom_line(data=cov.splines, aes(x=x.avg.pH, y=y.avg.pH, color="Understory"), linewidth=1.5) + 
  geom_line(data=pcq.splines, aes(x=x.avg.pH, y=y.avg.pH, color="Woody Vegetation"), linewidth=1.5, linetype="dashed") + 
  geom_line(data=bac.splines, aes(x=x.avg.pH, y=y.avg.pH, color="Bacterial"), linewidth=1.5) +
  geom_line(data=fun.splines, aes(x=x.avg.pH, y=y.avg.pH, color="Fungal"), linewidth=1.5) + 
   xlab("pH") + theme_classic() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(colour = "black", linewidth=1),
        axis.title.y=element_blank(), 
        legend.position = "NONE") + 
  ylim(0,3.4) + 
  ggtitle("") + 
 scale_color_manual(values=color)

elev.iS <- ggplot() + 
  geom_vline(aes(xintercept=0.698, color="Understory")) + 
  geom_vline(aes(xintercept=0.203, color="Woody Vegetation")) +
  geom_vline(aes(xintercept=0.699, color="Bacterial")) + 
  geom_vline(aes(xintercept=0.70, color="Fungal")) + 
  geom_line(data=cov.splines, aes(x=x.elevation, y=y.elevation, color="Understory"), linewidth=1.5, linetype="dashed") + 
  geom_line(data=pcq.splines, aes(x=x.elevation, y=y.elevation, color="Woody Vegetation"), linewidth=1.5, linetype="dashed") + 
  geom_line(data=bac.splines, aes(x=x.elevation, y=y.elevation, color="Bacterial"), linewidth=1.5, linetype="dashed") +
  geom_line(data=fun.splines, aes(x=x.elevation, y=y.elevation, color="Fungal"), linewidth=1.5, linetype="dashed") + 
  xlab("Elevation (m)") + 
  theme_classic() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(colour = "black", linewidth=1),
        axis.title.y=element_blank(), 
        legend.position = "NONE")  + 
  ylim(0,3.4) + 
   ggtitle("") + 
  scale_color_manual(values=color)
 
geo.iS <- ggplot() + 
   geom_line(data=cov.splines, aes(x=x.Geographic, y=y.Geographic, color="Understory"), linewidth=1.5) + 
   geom_line(data=pcq.splines, aes(x=x.Geographic, y=y.Geographic, color="Woody Vegetation"), linewidth=1.5) + 
   geom_line(data=bac.splines, aes(x=x.Geographic, y=y.Geographic, color="Bacterial"), linewidth=1.5) +
   geom_line(data=fun.splines, aes(x=x.Geographic, y=y.Geographic, color="Fungal"), linewidth=1.5) + 
   xlab("Geographic Distance (DD)") + theme_classic() + 
   theme(panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_rect(colour = "black", linewidth=1),
         axis.title.y=element_blank()) + 
  ylim(0,3.4) + 
  ggtitle("") + 
   scale_color_manual(values=color)

fig2 <- ggpubr::ggarrange(salt.iS, pH.iS, elev.iS, geo.iS, nrow=1, common.legend=TRUE, legend="bottom", labels=c("A","B","C","D"))

ggsave("Figure 2 GDMs.png", plot=fig2, width =9, height = 3, dpi=600)
```

### Mantel Tests for Matrix Correlation

```{r}
# Understory x woody vegetation
# Remove PAM 1 - no PCQ data for plot
cov.matrix.PAM1 <- cov.matrix %>% 
  rownames_to_column(var="plot.code") %>%
  filter(plot.code!="PAM01") %>% 
  arrange(plot.code) %>% 
  dplyr::select(-c(plot.code))

# Calculate relative abundance
cov.matrix.rel.PAM1 <- decostand(cov.matrix.PAM1, method = "total")

# Calculate dissimilarity matrix
cov.matrix.dist.PAM1 <- vegdist(cov.matrix.rel.PAM1, method = "bray")

# Format pcq matrix to remove rownames
pcq.matrix.PAM1 <- ba.matrix %>% 
  rownames_to_column(var="plot.code") %>%
  arrange(plot.code) %>% 
  dplyr::select(-c(plot.code))

# Calculate relative abundance
pcq.matrix.rel.PAM1 <- decostand(pcq.matrix.PAM1, method = "total")

# Calculate dissimilarity matrix
pcq.matrix.dist.PAM1 <- vegdist(pcq.matrix.rel.PAM1, method = "bray")


cov.pcq.mantel = mantel(cov.matrix.dist.PAM1, pcq.matrix.dist.PAM1, method = "spearman", permutations = 9999, na.rm = TRUE)

cov.pcq.mantel
# Mantel statistic r: 0.2335 
# Significance: 0.0019

# Fungi x understory
cov.fun.mantel = mantel(fun.matrix.dist, cov.matrix.dist, method = "spearman", permutations = 9999, na.rm = TRUE)

cov.fun.mantel
# Mantel statistic r: 0.4774 
# Significance: 1e-04

# Fungi x woody vegetation
# Remove PAM 1 - no PCQ data for plot
fun.otu.field.PAM1 <- fun.otu %>% 
  select(-c("X")) %>% 
  subset(plot.code!="PAM01") %>%
  arrange(plot.code) %>% 
  dplyr::select(-c(1:2))

# Calculate relative abundance
fun.matrix.rel.PAM1 <- decostand(fun.otu.field.PAM1, method = "total")

# Calculate dissimilarity matrix
fun.matrix.dist.PAM1 <- vegdist(fun.matrix.rel.PAM1, method = "bray")

pcq.fun.mantel = mantel(fun.matrix.dist.PAM1, pcq.matrix.dist.PAM1, method = "spearman", permutations = 9999, na.rm = TRUE)

pcq.fun.mantel
# Mantel statistic r: 0.1563 
# Significance: 0.0195 

# Bacteria x understory
cov.bac.mantel = mantel(bac.matrix.dist, cov.matrix.dist, method = "spearman", permutations = 9999, na.rm = TRUE)

cov.bac.mantel
# Mantel statistic r: 0.5715 
# Significance: 1e-04 

# Bacteria x woody vegetation
bac.otu.field.PAM1 <- bac.otu %>% 
  subset(plot.code!="PAM01") %>%
  arrange(plot.code) %>% 
  dplyr::select(-c(1:8))

## calculate relative abundance ##
bac.matrix.rel.PAM1 <- decostand(bac.otu.field.PAM1, method = "total")

## calculate dissimilarity matrix ##
bac.matrix.dist.PAM1 <- vegdist(bac.matrix.rel.PAM1, method = "bray")

pcq.bac.mantel = mantel(bac.matrix.dist.PAM1, pcq.matrix.dist.PAM1, method = "spearman", permutations = 9999, na.rm = TRUE)

pcq.bac.mantel
# Mantel statistic r: 0.1796 
# Significance: 0.0144

## Fungi x bacteria
bac.fun.mantel = mantel(bac.matrix.dist, fun.matrix.dist, method = "spearman", permutations = 9999, na.rm = TRUE)
bac.fun.mantel
# Mantel statistic r: 0.747 
# Significance: 1e-04 
```

##### FIGURE 4: Plotting Distance Matrices

```{r}
# create lists of distance value for each row/col combination (except Woody Vegetation) #
bac.dist.list <- bac.matrix.dist %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  filter(value!=0)

cov.dist.list <- cov.matrix.dist %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  subset(value!=0)

fun.dist.list <- fun.matrix.dist %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  subset(value!=0)

# combine lists #
bac.cov.dist.list <- left_join(bac.dist.list, cov.dist.list, by=c("row","col"))
fun.cov.dist.list <- left_join(fun.dist.list, cov.dist.list, by=c("row","col"))

# make separate lists for correlating with woody vegetation (PAM01 omitted) #
# PAM01 was formerly labeled PAM06 #

bac.dist.list.PAM1 <- bac.matrix.dist.PAM1 %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  subset(value!=0)

fun.dist.list.PAM1 <- fun.matrix.dist.PAM1 %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  subset(value!=0)

pcq.dist.list <- pcq.matrix.dist.PAM1 %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  subset(value!=0)

cov.dist.list.PAM1 <- cov.matrix.dist.PAM1 %>% 
  as.matrix() %>% 
  melt(varnames = c("row", "col")) %>% 
  subset(value!=0)

# combine lists #
bac.pcq.dist.list <- left_join(bac.dist.list.PAM1, pcq.dist.list, by=c("row","col"))
fun.pcq.dist.list <- left_join(fun.dist.list.PAM1, pcq.dist.list, by=c("row","col"))
cov.pcq.dist.list <- left_join(cov.dist.list.PAM1, pcq.dist.list, by=c("row","col"))
bac.fun.dist.list <- left_join(fun.dist.list, bac.dist.list, by=c("row","col"))


# plot distance values for each pairing #
bac.cov.mant <- ggplot() + 
  geom_point(data=bac.cov.dist.list, aes(x=value.y, y=value.x, fill="Understory"), size=2, shape=21, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(data=bac.cov.dist.list, aes(x=value.y, y=value.x, fill="Understory"), color="black", fullrange=TRUE, method=lm, linewidth=0.5) + 
  ylab("Bacterial Distance") + #xlab("Understory Distance") + 
  geom_label(aes(x = 0.2, y = 0.2, label = "Mant stat=0.57\nP=<0.001"), size=3) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  theme(legend.position="NONE") +
  ggtitle("") + xlab("") + 
  ylim(0,1) + xlim(0,1)

bac.pcq.mant <- ggplot() + 
  geom_point(data=bac.pcq.dist.list, aes(x=value.y, y=value.x, fill="Woody Vegetation"), size=2, shape=21, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(data=bac.pcq.dist.list, aes(x=value.y, y=value.x, fill="Woody Vegetation"), color="black", fullrange=TRUE, method=lm, linewidth=0.5) + 
  geom_label(aes(x = 0.2, y = 0.2, label = "Mant stat=0.18\nP=0.014"), size=3) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  theme(legend.position="NONE") +
  ggtitle("") + 
  ylab("") + xlab("") + 
  ylim(0,1) + xlim(0,1)
 
fun.cov.mant <- ggplot() + 
  geom_point(data=fun.cov.dist.list, aes(x=value.y, y=value.x, fill="Understory"), size=2, shape=21, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(data=fun.cov.dist.list, aes(x=value.y, y=value.x, fill="Understory"), color="black", fullrange=TRUE, method=lm, linewidth=0.5) + 
  ylab("Fungal Distance") + xlab(" ") + 
  geom_label(aes(x = 0.2, y = 0.2, label = "Mant stat=0.48\nP=<0.001"), size=3) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  theme(legend.position = "NONE") + 
  ggtitle("") + 
  ylim(0,1) + xlim(0,1)

fun.pcq.mant <- ggplot() + 
  geom_point(data=fun.pcq.dist.list, aes(x=value.y, y=value.x, fill="Woody Vegetation"), size=2, shape=21, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(data=fun.pcq.dist.list, aes(x=value.y, y=value.x, fill="Woody Vegetation"), color="black", fullrange=TRUE, method=lm, linewidth=0.5) + 
  xlab("Woody Distance") + 
  geom_label(aes(x = 0.2, y = 0.2, label = "Mant stat=0.16\nP=0.020"), size=3) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  theme(legend.position = "NONE") + 
  ylab(" ") + 
  ggtitle("") + 
  ylim(0,1) + xlim(0,1)

pcq.cov.mant <- ggplot() + 
  geom_point(data=cov.pcq.dist.list, aes(x=value.x, y=value.y, fill="Understory"), size=2, shape=21, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(data=cov.pcq.dist.list, aes(x=value.x, y=value.y, fill="Understory"), color="black", fullrange=TRUE, method=lm, linewidth=0.5) + 
  ylab("Woody Distance") + xlab("Understory Distance") + 
  geom_label(aes(x = 0.2, y = 0.2, label = "Mant stat=0.23\nP=0.002"), size=3) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  theme(legend.position = "NONE") + 
  ggtitle("") + 
  ylim(0,1) + xlim(0,1)

microbe.mant <- ggplot() + 
  geom_point(data=bac.fun.dist.list, aes(x=value.x, y=value.y, fill="Fungal"), size=2, shape=21, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(data=bac.fun.dist.list, aes(x=value.x, y=value.y, fill="Fungal"), color="black", fullrange=TRUE, method=lm, linewidth=0.5) + 
  xlab("Fungal Distance") + 
  ylab("Bacterial Distance") +
  geom_label(aes(x = 0.2, y = 0.2, label = "Mant stat=0.75\nP=<0.001"), size=3) + 
  scale_color_manual(values=color) + 
  scale_fill_manual(values=color) + 
  theme(legend.position = "NONE") + 
  ggtitle("") + 
  ylim(0,1) + xlim(0,1)

fig5 <- ggpubr::ggarrange(bac.cov.mant, bac.pcq.mant, fun.cov.mant, fun.pcq.mant, pcq.cov.mant, microbe.mant, 
                  labels=c("A","B","C","D","E","F"), nrow=3, ncol=2)

ggsave("Figure 4 Mantel.png", plot=fig5, width=6, height=7, dpi=600)
```

### TITAN Analyses

```{r}
# Extract plot SWISLR driver values
ppt.titan <- env %>% 
  arrange(plot.code) %>%
  select(ppt)

pH.titan <- env %>% 
  arrange(plot.code) %>%
  select(avg.pH)

elev.titan <- env %>% 
  arrange(plot.code) %>%
  select(elevation)
```


##### Fungi

```{r}
# Extract OTUs with >3 occurences
fun.otu.pa <- decostand(fun.otu %>% arrange(plot.code) %>% select(-c(1:2)), method = "pa") 
fun.otu.pa <- fun.otu.pa[colSums(fun.otu.pa)>3]
fun.common_cols <- intersect(names(fun.otu.pa), names(fun.otu))

fun.otu.titan <- fun.otu %>% 
  select(all_of(fun.common_cols))

# Run TITAN analyses
set.seed(1)
fun.ppt.titan <- titan(ppt.titan, fun.otu.titan, nBoot=1000)

set.seed(1)
fun.pH.titan <- titan(pH.titan, fun.otu.titan, nBoot=1000)

set.seed(1)
fun.elev.titan <- titan(elev.titan, fun.otu.titan, nBoot=1000)

# View change point bootstrap quantiles
fun.ppt.titan$sumz.cp
fun.pH.titan$sumz.cp
fun.elev.titan$sumz.cp

## extract and save output ##
fun.ppt.titan.df <- fun.ppt.titan$sppmax %>%
  as.data.frame() %>%
  select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>% 
  rownames_to_column(var="OTU") %>% 
  left_join(fun.tax, by="OTU")

fun.pH.titan.df <- fun.pH.titan$sppmax %>%
  as.data.frame() %>%
  select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>% 
  rownames_to_column(var="OTU") %>% 
  left_join(fun.tax, by="OTU")

fun.elev.titan.df <- fun.elev.titan$sppmax %>%
  as.data.frame() %>%
  select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>% 
  rownames_to_column(var="OTU") %>% 
  left_join(fun.tax, by="OTU")

# Visualize taxa
plot_taxa_ridges(fun.ppt.titan) 
plot_taxa_ridges(fun.pH.titan) 
plot_taxa_ridges(fun.elev.titan) 
```

##### Bacteria

```{r}
# Extract OTUs with >3 occurences
bac.otu.pa <- decostand(bac.otu, method = "pa") 
bac.otu.pa <- bac.otu.pa[colSums(bac.otu.pa)>3]
common_cols <- intersect(names(bac.otu.pa), names(bac.otu))

bac.otu.titan <- bac.otu %>% 
  select(all_of(common_cols))

# Run TITAN analyses
set.seed(1)
bac.ppt.titan <- titan(ppt.titan, bac.otu.titan, nBoot=1000)

set.seed(1)
bac.pH.titan <- titan(pH.titan, bac.otu.titan, nBoot=1000)

set.seed(1)
bac.elev.titan <- titan(elev.titan, bac.otu.titan, nBoot=1000)

# View change point bootstrap quantiles
bac.ppt.titan$sumz.cp
bac.pH.titan$sumz.cp
bac.elev.titan$sumz.cp


# Extract and save output
bac.ppt.titan.df <- bac.ppt.titan$sppmax %>%
  as.data.frame() %>%
  select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>% 
  rownames_to_column(var="OTU") %>% 
  left_join(bac.tax, by="OTU")

bac.pH.titan.df <- bac.pH.titan$sppmax %>%
  as.data.frame() %>%
  select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>% 
  rownames_to_column(var="OTU") %>% 
  left_join(bac.tax, by="OTU")

bac.elev.titan.df <- bac.elev.titan$sppmax %>%
  as.data.frame() %>%
  select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>% 
  rownames_to_column(var="OTU") %>% 
  left_join(bac.tax, by="OTU")

# Visualize taxa
plot_taxa_ridges(bac.ppt.titan) 
plot_taxa_ridges(bac.pH.titan) 
plot_taxa_ridges(bac.elev.titan) 
```

##### Figure 5 TITAN 

```{r}
# Fungal change point and density plots
fun.pH.titan.plot <- plot_sumz_density(fun.pH.titan, 
                                       sumz=FALSE, 
                                       col1="#FF7CAC", col2="#ffbad3", 
                                       xlabel=NULL, y2label=NULL)

fun.ppt.titan.plot <- plot_sumz_density(fun.ppt.titan, 
                                        sumz=FALSE, 
                                        col1="#FF7CAC", col2="#ffbad3", 
                                        xlabel=NULL, y2label=NULL)

fun.elev.titan.plot <- plot_sumz_density(fun.elev.titan, 
                                         sumz=FALSE, 
                                         col1="#FF7CAC", col2="#ffbad3", 
                                         xlabel=NULL, y2label=NULL)

# Bacterial change point and density plots
bac.pH.titan.plot <- plot_sumz_density(bac.pH.titan, 
                                       sumz=FALSE, 
                                       col1="#2949C5", col2="#758be0", 
                                       xlabel=NULL, y2label=NULL)

bac.ppt.titan.plot <- plot_sumz_density(bac.ppt.titan, 
                                        sumz=FALSE, 
                                        col1="#2949C5", col2="#758be0", 
                                        xlabel=NULL, y2label=NULL)

bac.elev.titan.plot <- plot_sumz_density(bac.elev.titan, 
                                         sumz=FALSE, 
                                         col1="#2949C5", col2="#758be0", 
                                         xlabel=NULL, y2label=NULL)

ggarrange(fun.ppt.titan.plot, fun.pH.titan.plot, fun.elev.titan.plot, 
          bac.ppt.titan.plot, bac.pH.titan.plot, bac.elev.titan.plot, 
          labels=c("A", "B","C","D","E","F")) 

ggsave("Figure 5 TITAN.png", width = 6, height = 4, dpi=600)

```

### Supplemental Figures

##### S Figure 1: Vegetation Abundance

```{r}
MAS.data <- veg.ba.sum %>% filter(transect=="MAS")
PPR.data <- veg.ba.sum %>% filter(transect=="PPR")
PAM.data <- veg.ba.sum %>% filter(transect=="PAM")

MAS.under <- ggplot(MAS.data) + 
  geom_area(aes(x=distance, y=herb.sum, fill="Understory")) +
  theme_classic() + ylab("% cover") +  ggtitle("MAS") + 
  scale_fill_manual(values=color) + 
  ylim(0,100) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PPR.under <- ggplot(PPR.data) + 
  geom_area(aes(x=distance, y=herb.sum, fill="Understory")) +
  theme_classic() + ylab("% cover") +  ggtitle("PPR") + 
  scale_fill_manual(values=color) + 
  ylim(0,100) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PAM.under <- ggplot(PAM.data) + 
  geom_area(aes(x=distance, y=herb.sum, fill="Understory")) +
  theme_classic() + ylab("% cover") +  ggtitle("PAM") + 
  ylim(0,100) + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

MAS.shrub <- ggplot(MAS.data) + 
  geom_area(aes(x=distance, y=shrub.ba, fill="Shrub")) +
  theme_classic() + ylab("basal area") +
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PPR.shrub <- ggplot(PPR.data) + 
  geom_area(aes(x=distance, y=shrub.ba, fill="Shrub")) +
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PAM.shrub <- ggplot(PAM.data) + 
  geom_area(aes(x=distance, y=shrub.ba, fill="Shrub")) +
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

MAS.tree <- ggplot(MAS.data) + 
  geom_area(aes(x=distance, y=tree.ba, fill="Woody Vegetation")) +
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PPR.tree <- ggplot(PPR.data) + 
  geom_area(aes(x=distance, y=tree.ba, fill="Woody Vegetation")) +
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PAM.tree <- ggplot(PAM.data) + 
  geom_area(aes(x=distance, y=tree.ba, fill="Woody Vegetation")) +
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

MAS.dead <- ggplot(MAS.data) + 
  geom_area(aes(x=distance, y=dead.sum, fill="Dead Tree")) + 
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PPR.dead <- ggplot(PPR.data) + 
  geom_area(aes(x=distance, y=dead.sum, fill="Dead Tree")) + 
  theme_classic() + ylab("basal area") + 
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

PAM.dead <- ggplot(PAM.data) + 
  geom_area(aes(x=distance, y=dead.sum, fill="Dead Tree")) + 
  theme_classic() +
  scale_fill_manual(values=color) + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  geom_point(aes(x=distance, y=0), color="black")

MAS.ppt <- ggplot(MAS.data) + 
  geom_smooth(aes(x=distance, y=ppt), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=ppt, fill="black", alpha=0.5), size=3) + 
  theme_classic() + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(0,13.5)

PPR.ppt <- ggplot(PPR.data) + 
  geom_smooth(aes(x=distance, y=ppt), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=ppt, fill="black", alpha=0.5), size=3) + 
  theme_classic() +
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(0,13.5)

PAM.ppt <- ggplot(PAM.data) + 
  geom_smooth(aes(x=distance, y=ppt), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=ppt, fill="black", alpha=0.5), size=3) + 
  theme_classic() + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(0,13.5)

MAS.pH <- ggplot(MAS.data) + 
  geom_smooth(aes(x=distance, y=avg.pH), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=avg.pH, fill="black", alpha=0.5), size=3) + 
  theme_classic() + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(3.7,6.1)

PPR.pH <- ggplot(PPR.data) + 
  geom_smooth(aes(x=distance, y=avg.pH), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=avg.pH, fill="black", alpha=0.5), size=3) + 
  theme_classic() +
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(3.7,6.1)

PAM.pH <- ggplot(PAM.data) + 
  geom_smooth(aes(x=distance, y=avg.pH), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=avg.pH, fill="black", alpha=0.5), size=3) + 
  theme_classic() + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(3.7,6.1)

MAS.elev <- ggplot(MAS.data) +
  geom_smooth(aes(x=distance, y=elevation), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=elevation, fill="black", alpha=0.5), size=3) + 
  theme_classic() + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(0.2, 0.7)

PPR.elev <- ggplot(PPR.data) + 
  geom_smooth(aes(x=distance, y=elevation), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=elevation, fill="black", alpha=0.5), size=3) + 
  theme_classic() +
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(0.2, 0.7)

PAM.elev <- ggplot(PAM.data) + 
  geom_smooth(aes(x=distance, y=elevation), method="lm", se=FALSE, color="black", size=0.5) + 
  geom_point(aes(x=distance, y=elevation, fill="black", alpha=0.5), size=3) + 
  theme_classic() + 
  theme(plot.title=element_text(size=16), 
        legend.position = "NONE",
        axis.title.x = element_blank(),
        axis.title.y = element_blank()) + 
  ylim(0.2, 0.7)
 

sfig1 <- egg::ggarrange(MAS.under, PPR.under, PAM.under,
               MAS.shrub, PPR.shrub, PAM.shrub, 
               MAS.tree, PPR.tree, PAM.tree, 
               MAS.dead, PPR.dead, PAM.dead, 
               MAS.ppt, PPR.ppt, PAM.ppt,
               MAS.pH, PPR.pH, PAM.pH,
               MAS.elev, PPR.elev, PAM.elev,
               ncol=3, nrow=7,bottom="distance from marsh (m)")

ggsave(plot=sfig1,"SI Figure 1 Vegetation.png",  width = 6.25, height = 8.5, dpi=600)
```

#### S Figure 2: Variable correlations

```{r}
FXS <- 21
FXC <- "black"
FXF <- "black"

summary(lm(ppt~elevation, data=env))
summary(lm(avg.pH~elevation, data=env)) 
summary(lm(avg.pH~ppt, data=env)) 
summary(lm(percent.sm~elevation, data=env)) 
summary(lm(percent.sm~ppt, data=env)) 

elev.ppt <- ggplot(env, aes(x=elevation, y=ppt)) + 
  geom_point(size=4, shape=FXS, color=FXC, fill=FXF, alpha=0.5) + 
  theme_classic() + 
  ylab("Salinity (ppt)") + 
  xlab("elevation (m)") + 
  theme(axis.title.x = element_blank()) + 
  geom_label(aes(x = 0.6, y = 12.5, label = "R: -0.048"), size=3) + 
  ggtitle("A")

elev.pH <- ggplot(env, aes(x=elevation, y=avg.pH)) + 
  geom_point(size=4, shape=FXS, color=FXC, fill=FXF, alpha=0.5) + 
  theme_classic() + 
  ylab("pH")  + 
  ylim(4,8) + 
  theme(axis.title.x = element_blank())  + 
  geom_label(aes(x = 0.6, y = 7.5, label = "R: -0.248"), size=3) + 
  ggtitle("B")

ppt.pH <- ggplot(env, aes(x=ppt, y=avg.pH)) + 
  geom_point(size=4, shape=FXS, color=FXC, fill=FXF, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(method="glm", 
              method.args=list(family=gaussian(link="log")), 
              fill=FXC,  color=FXC,
              linewidth=0.5, alpha=0.25) + 
  ylim(4,8) + 
  theme(axis.title.y = element_blank(), axis.title.x = element_blank())  + 
  geom_label(aes(x = 11, y = 7.5, label = "R: 0.647"), size=3) + 
  ggtitle("C")


elev.sm <- ggplot(env, aes(x=elevation, y=percent.sm)) + 
  geom_point(size=4, shape=FXS, color=FXC, fill=FXF, alpha=0.5) + 
  theme_classic() + 
  geom_smooth(method="glm", 
              method.args=list(family=gaussian(link="log")), 
              fill=FXC,  color=FXC,
              linewidth=0.5, alpha=0.25) + 
  ylab("Soil moisture (%)") + 
  xlab("elevation (m)")  + 
  ylim(38,110) +
  geom_label(aes(x = 0.6, y = 100, label = "R: -0.570"), size=3) + 
  ggtitle("D")


ppt.sm <- ggplot(env, aes(x=ppt, y=percent.sm)) + 
  geom_point(size=4, shape=FXS, color=FXC, fill=FXF, alpha=0.5) + 
  theme_classic() + 
  xlab("Salinity (ppt)") + 
  ylim(38,110) +
  theme(axis.title.y = element_blank())  + 
  geom_label(aes(x = 11, y = 100, label = "R: 0.222"), size=3) + 
  ggtitle("E")

spacer <- ggplot() + theme_void()

sfig2 <- egg::ggarrange(elev.ppt, spacer, elev.pH, ppt.pH, elev.sm, ppt.sm, 
          nrow=3, ncol=2)

ggsave(plot=sfig2, "SI Figure 2 Variable Correlation.png", width = 4.5, height = 5.5, dpi=600)
```

